Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-08 paper/abstract.tex
various little changes

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first-stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor and also
facilitates modular expression of staged programs.

>>>>

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first-stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that
improve code efficiency by up to a linear factor and also
facilitates modular expression of staged programs.

### 2016-01-08 paper/intro.tex
various little changes

\input{figures/quickselect}
Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this scenario, a multivariate function is applied many times with
one argument fixed, so it is advantageous to find the computations in
$f$ that depend on $x$ and then stage program execution so they are
performed only once at the beginning of execution.

@@

One common precomputation technique is partial
evaluation\,\cite{futamura71,jones96}, which relies on dynamic
compilation to specialize functions to known argument values. Going
back to our example, if $f$ specializes to a particular $v$, written
$f_v$, such that $f(v,y) = f_v(y)$, then $F$ can be specialized to $v$
as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$
which do not depend on the second argument.

>>>>

\input{figures/quickselect}
Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
Observe that this would be wasteful if $f(x,y_i)$ does a significant 
amount of work that does not depend on $y_i$.
In such a case, it would be advantageous to find the computations in
$f$ that depend only on $x$ and then stage execution of $f$ so they are
performed only once at the beginning of $F$.

@@

One common precomputation technique is partial
evaluation\,\cite{futamura71,jones96}, which relies on dynamic
compilation to specialize functions to known argument values. Going
back to our example, if $f$ specializes to a particular $v$, written
$f_v$, such that $f(v,y) = f_v(y)$, then $F$ can be specialized to $v$
as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
This eliminates the need to compute $m$ times those parts of $f(v,-)$
which do not depend on the second argument.

### 2016-01-08 paper/overview.tex
Carlo's comments
various little changes

Suppose that we wish to perform a series of order statistics queries
on a list~\texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

A first cut at a solution might be to first sort \texttt{l} into
\texttt{s} and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l} and then simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. For a sufficiently balanced tree, this approach has 
(expected) runtime $\Theta(n\log{n} + m\log{n})$, 
a nearly linear-time improvement in $m$.

@@

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code. Additionally, $\texttt{gr}$ (\emph{ground}) annotations 
mark certain first stage components as being purely first-stage, rather than mixed-stage.
%
We also use a function 
\lstinline{1`hold : ^int -> $`2`int`}, 
\begin{comment}$ this comment block is here for syntax highlighting reasons\end{comment}
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

>>>>

Suppose that we wish to perform a series of order statistics queries
on a list~\texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and an integer
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k} (i.e. the $k$th-largest element).
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input list, with size $n$, is uniformly
randomly ordered (which can be achieved by pre-permuting it), \texttt{qs}
runs in expected $\Theta(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $\Theta(n \cdot m)$ time.

@@

A first cut at a solution might be to first sort \texttt{l} into
a sorted list \texttt{s} and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $\Theta(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l} and then simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. For a sufficiently balanced tree, 
which one would get with a random choice of pivot, this approach has 
expected runtime $\Theta(n\log{n} + m\log{n})$, 
a nearly linear-time improvement in $m$.

@@

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code. Additionally, $\texttt{gr}$ (\emph{ground}) annotations 
mark certain first-stage components as being purely first-stage, rather than mixed-stage.
%
We also use a function 
\lstinline{1`hold : ^int -> $`2`int`}, 
\begin{comment}$ this comment block is here for syntax highlighting reasons\end{comment}
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.
